Forecasting interrupted time series

### 2024-05-17 response_letter/response_letter_JORS.qmd

_Comment 1: The topic is very interesting and very timely, in the wake of the Covid -19 related disruptions in supply chains, changes in consumer preferences, and inflationary environment. The time series patterns and causal relationships that were seen before the pandemic did not apply and whether the patterns observed before the pandemic will continue to be relevant remains to be evaluated on a case by case basis._

**Response:** We are delighted to hear that you appreciate the timeliness and importance of the paper. We agree that the topic is particularly relevant given the COVID-19-related disruptions in supply chains and various other areas. We also agree that evaluating whether pre-pandemic time series patterns and causal relationships remain applicable is a critical challenge that requires careful case-by-case analysis. That is why, in this paper, we propose various strategies rather than a single approach to address different situations and cases effectively.

We have added the following paragraph in the revised manuscript to address this comment:

\textcolor{teal}{The handling of time series interruptions is a complex and multifaceted issue, and the choice of the most appropriate strategy depends on the specific characteristics of the data and the nature of the interruption. Therefore, in this paper, we propose multiple strategies rather than a single approach to effectively address different situations and cases.}

@@

\textcolor{teal}{Moreover, in determining which strategy to apply in each case, it is important to think through what is being assumed, and what data are being used. This is a qualitative and subjective process, and cannot simply be tackled with a quantitative comparison of forecast accuracy. The strategies we discuss have different ways of handling the disrupted period, and many of them do not seek to provide accurate forecasts during the disruption. For example, the "set to missing" strategy ignores the disrupted period, while the "estimate what might have been" strategy replaces the observations during the disruption with interpolations. In both cases, it does not make sense to evaluate the performance of the method during the disruption. Instead, the analyst needs to decide what approach makes the most sense for the application at hand.}

_Comment 4. If the authors focus on the interrupted time series for a certain event, some exogeneous variables are encouraged for potential better performance._

**Response:** Thank you for your suggestion. We agree that incorporating exogenous variables related to specific events could enhance the performance of forecasting models for interrupted time series. We have acknowledged this by adding the following paragraph to the revised manuscript:

\textcolor{teal}{When forecasting interrupted time series data, exogenous variables associated with specific events should be identified and incorporated where possible, regardless of which approach is employed, since they may improve forecasting model performance.}

@@

- The data sets are described in Section 3.

- Insights are provided in Sections 3 and 4.

_Comment 3. The figures presented in the manuscripts require detailed explanation. Authors are required to explain the given figures in detail one by one._

**Response:** Thank you for your feedback. We have provide a detailed explanation of each figure in the manuscript, discussing them one by one to ensure clarity. We have now added some of the same information in the captions of each figure to aid readability.

### 2024-05-17 fits.qmd
Final changes to revised paper and response

Time series forecasting models use historical data to estimate future values [@fildes2008forecasting] based on consistent patterns observed in the past. However, time series are sometimes disrupted by unusual events that jeopardise the regularity of the time series pattern. This can be defined as an *interrupted time series*. Time series data may be disrupted by a variety of reasons, ranging from temporary policy changes to natural disasters. Outbreaks, epidemics, and pandemics, for example, may have a considerable impact on population behavior, causing discrepancies and interruptions of data in a variety of industries such as healthcare, pharmaceuticals, transportation, retail, tourism, and traffic management. Similarly, transportation-related data may be interrupted due to equipment failures, such as malfunctions or breakdowns in vehicles, aircraft, or rail systems, resulting in missing or inaccurate data points. Furthermore, cybersecurity breaches data quality by possibly skewing readings in datasets relating to financial transactions, network traffic, and user activity, complicating time series analysis. Supply chain disruptions induced by factors such as strikes or shortages cause irregularities in sales, inventory levels, production, and delivery, often resulting in data gaps. \textcolor{teal}{The war in Ukraine caused disruptions in sunflower oil and wheat supply chains, which were most likely driven by social media and people's recent memories of COVID-19 toilet paper shortage. A power outage in Eastern Germany halted production at a Tesla facility and cut off electricity to a large retailer distribution centre, disrupting supplies to around 500 stores in and around Berlin for several weeks. Recent strikes in various forms of transport, including air and rail in Germany, France, and the United Kingdom, have had an impact on passenger flow and retail sales at airports and train stations. These issues highlight the significance of robust forecasting systems in addressing such disruptions effectively.}

@@

It is important to note that in this study, we consider the problem of forecasting interrupted time series both during and after the event. This is a separate problem from change point detection [@aue2013structural;@truong2020selective] and anomaly detection [@talagala2020anomaly]. In the situations we investigate, we know that a change has occurred which has caused some unusual observations, and we want to forecast what will happen next. Change point detection refers to determining *when* a change occurred [@blazquez2021review], and anomaly detection aims to identify unusual observations with the intention of either minimising their value, or paying particular attention to them, and doing a root-cause investigation. Our study focus also differs from *intervention modeling*, sometimes referred to as *interrupted time series analysis and modelling* [@mcdowall2019interrupted], which is used when we have data about an outcome over time and the aim is to assess the impact of an intervention, policy, or program implemented at a specific time point. We are not trying to measure the impact of the interruption; rather we are trying to produce sensible forecasts during and after the interruption.

@@

In this section, we describe several possible strategies to handle interruptions when forecasting time series data. \textcolor{teal}{We note that the assessment of whether pre-pandemic time series patterns and underlying relationships remain applicable is a critical issue that requires careful case-by-case analysis. Therefore, in this paper, we propose multiple strategies rather than a single approach to effectively address different situations and cases.}

@@

\textcolor{teal}{The disadvantage of smoothing and other highly adaptive techniques is that they need time to adapt after a disturbance and can eventually end up chasing noise. The latter is an intriguing issue since individuals often prefer adaptability while failing to recognise that it does not necessarily improve forecasting, ignoring the presence of noise in the data.}

@@

This has the advantage of retaining the memory of the past, and so the seasonal patterns and other dynamics will be retained. This allows the change period to be effectively modelled, provided the intervention variables are chosen well. \textcolor{teal}{Intervention models are useful because they enable the forecasting of similar future disruptions, but distinguishing between a disruption and a general driving factor, such as a promotion or a seasonal storm, might become subjective.} However, the model will assume that the post-interruption period will be similar to the pre-interruption period, and so the prediction intervals may be too narrow, especially if there is a lasting effect beyond the end of the interruption.

@@

An alternative would be to take the model estimated under the previous solution (setting the observations during the disruption to missing), and use it to estimate what might have been during the disruption. Then the model can be re-estimated using the adjusted data. Note that forecasts obtained in this way during the disruption will not be true forecasts because they will have used data from the future in computing the adjusted data during the disruption. But post-disruption forecasts will be true forecasts and should be almost the same as those obtained using the previous solution.

@@

```{r}
#| label: fig-tourism-plot1
#| fig-cap: "Short-term visitor arrivals to Australia (monthly): Jan 2000 -- May 2023."
#| fig-height: 3
tar_read(tourism_plot1)
```

@@

```{r}
#| label: fig-tsol1-plot
#| fig-height: 8
#| fig-pos: "!b"
#| fig-cap: "**Highly adaptive models**. Forecasts from ETS and ARIMA models. Neither works particularly well for disruptions of this magnitude."
tar_read(tsol1_plot)
```

@@

```{r}
#| label: fig-tsol2-plot
#| fig-height: 8
#| fig-pos: "!b"
#| fig-cap: "**Intervention model**. We use a level shift from March 2020 to October 2022, and a ramp from October 2021 to October 2022. After October 2022, all intervention variables are set to zero."
tar_read(tsol2_plot)
```

@@

```{r}
#| label: fig-tsol3-plot
#| fig-height: 8
#| fig-pos: "!b"
#| fig-cap: "**Set to missing**. The period from March 2020 to October 2022 is set to missing. So the first three years of forecasts show what might have been. The fourth set of forecasts uses observations from November and December 2022, and so the forecasts have been adjusted downwards."
tar_read(tsol3_plot)
```

@@

```{r}
#| label: fig-tsol4-plot
#| fig-height: 8
#| fig-pos: "!b"
#| fig-cap: "**Estimate what might have been**. We replace the observations from March 2020 to October 2022 with the average of the same month in the three years prior to March 2020."
tar_read(tsol4_plot)
```

@@

\FloatBarrier

@@

```{r}
#| label: fig-walkers
#| fig-height: 3.7
#| fig-cap: "Pedestrian traffic in Melbourne, Australia, from 2018 to 2023. The gray shaded regions indicate periods of lockdown due to COVID-19."
tar_read(walkers_plot)
```

@@

```{r}
#| label: fig-walkers-plot1
#| fig-height: 3
#| fig-cap: "**Highly adaptive model**: the forecasts do not take into account the lockdown periods, but the ARIMA model is largely able to adjust to the changing level of the series. Forecasts are produced on a rolling basis each week, with a 7-day forecast horizon in each case."
tar_read(walkers_plot1)
```

@@

@@

```{r}
#| label: fig-walkers-plot2
#| fig-height: 3
#| fig-cap: "**Intervention model**: the forecasts are based on an ARIMA model with a dummy variable indicating when the lockdown periods occurred. Many people were in self-imposed lockdown before the start of the first lockdown period, and were reluctant to return to the city after it ended, resulting in poor forecasts on either side of the first lockdown. The model was able to adjust to the changing level of the series during subsequent lockdowns."
tar_read(walkers_plot2)
```

```{r}
#| label: fig-walkers-plot3
#| fig-height: 3
#| fig-cap: "**Set to missing**: the lockdown periods are set to missing, and the ARIMA model uses the remaining data to produce forecasts. Thus, the forecasts are based on only previous non-lockdown data. Naturally, the forecasts during the lockdown periods do not reflect reality, but they can be interpreted as what might have been."
tar_read(walkers_plot3)
```

```{r}
#| label: fig-walkers-plot4
#| fig-height: 3
#| fig-cap: "**Estimate what might have been**: the lockdown periods are replaced with estimated counts based on an ARIMA model applied to the remaining data. Then a new ARIMA model is fitted to the whole data set, and used to produce forecasts. As a result, the forecasts during the lockdown periods are not true forecasts, as they use non-lockdown data from the future."
tar_read(walkers_plot4)
```

```{r}
#| label: fig-walkers-ensemble-plot
#| fig-height: 3
#| fig-cap: "**Ensemble forecasts** based on the first two solutions shown in Figures [-@fig-walkers-plot1] and [-@fig-walkers-plot2]. Other than having some difficulty around the period of the first lockdown, the results are relatively good."
tar_read(walkers_ensemble_plot)
```

@@

\textcolor{teal}{We should also note that when forecasting interrupted time series data, exogenous variables associated with specific events should be identified and incorporated, since they may improve forecasting model performance.}

@@

We would like to thank Dr. Stephan Kolassa for sharing his thoughts and providing valuable feedback on the first version of our paper. His voluntary feedback and recommendations have helped us to refine and improve our work.

>>>>

Time series forecasting models use historical data to estimate future values [@fildes2008forecasting] based on consistent patterns observed in the past. However, time series are sometimes disrupted by unusual events that jeopardise the regularity of the time series pattern. This can be defined as an *interrupted time series*. Time series data may be disrupted by a variety of reasons, ranging from temporary policy changes to natural disasters. Outbreaks, epidemics, and pandemics, for example, may have a considerable impact on population behavior, causing discrepancies and interruptions of data in a variety of industries such as healthcare, pharmaceuticals, transportation, retail, tourism, and traffic management. Similarly, transportation-related data may be interrupted due to equipment failures, such as malfunctions or breakdowns in vehicles, aircraft, or rail systems, resulting in missing or inaccurate data points. Furthermore, cybersecurity breaches data quality by possibly skewing readings in datasets relating to financial transactions, network traffic, and user activity, complicating time series analysis. Supply chain disruptions induced by factors such as strikes or shortages cause irregularities in sales, inventory levels, production, and delivery, often resulting in data gaps. \textcolor{teal}{For example, the war in Ukraine caused disruptions in sunflower oil and wheat supply chains, which were most likely driven by social media and people's recent memories of the COVID-19 toilet paper shortage. A power outage in Eastern Germany halted production at a Tesla facility and cut off electricity to a large retailer distribution centre, disrupting supplies to around 500 stores in and around Berlin for several weeks. Recent strikes in various forms of transport, including air and rail in Germany, France, and the United Kingdom, have had an impact on passenger flow and retail sales at airports and train stations. These issues highlight the significance of robust forecasting systems in addressing such disruptions effectively.}

@@

We consider the problem of forecasting interrupted time series both during and after the event. This is a separate problem from change point detection [@aue2013structural;@truong2020selective] and anomaly detection [@talagala2020anomaly]. In the situations we investigate, we know that a change has occurred which has caused some unusual observations, and we want to forecast what will happen next. Change point detection refers to determining *when* a change occurred [@blazquez2021review], and anomaly detection aims to identify unusual observations with the intention of either minimising their value, or paying particular attention to them, and doing a root-cause investigation. Our study focus also differs from *intervention modeling*, sometimes referred to as *interrupted time series analysis and modelling* [@mcdowall2019interrupted], which is used when we have data about an outcome over time and the aim is to assess the impact of an intervention, policy, or program implemented at a specific time point. We are not trying to measure the impact of the interruption; rather we are trying to produce sensible forecasts during and after the interruption.

@@

In this section, we describe several possible strategies to handle interruptions when forecasting time series data. \textcolor{teal}{The handling of time series interruptions is a complex and multifaceted issue, and the choice of the most appropriate strategy depends on the specific characteristics of the data and the nature of the interruption. Therefore, in this paper, we propose multiple strategies rather than a single approach to effectively address different situations and cases.}

@@

\textcolor{teal}{The disadvantage of smoothing and other highly adaptive techniques is that they need time to adapt after a disturbance and can eventually end up chasing noise. The latter is an intriguing issue since individuals often prefer adaptability while failing to recognise that it does not necessarily improve forecasting.}

@@

This has the advantage of retaining the memory of the past, and so the seasonal patterns and other dynamics will be retained. This allows the change period to be effectively modelled, provided the intervention variables are chosen well. \textcolor{teal}{Intervention models are useful because they enable the forecasting of similar future disruptions, but distinguishing between a disruption and a general driving factor, such as a promotion or a seasonal storm, may become subjective.} However, the model will assume that the post-interruption period will be similar to the pre-interruption period, and so the prediction intervals may be too narrow, especially if there is a lasting effect beyond the end of the interruption.

@@

An alternative would be to take the model estimated under the previous solution (setting the observations during the disruption to missing), and use it to estimate what might have been during the disruption. Then the model can be re-estimated using the adjusted data. Forecasts obtained in this way during the disruption will not be true forecasts because they will have used data from the future in computing the adjusted data during the disruption. But post-disruption forecasts will be true forecasts and should be almost the same as those obtained using the previous solution.

@@

```{r}
#| label: fig-tourism-plot1
#| fig-cap: "Short-term visitor arrivals to Australia (monthly): Jan 2000 -- May 2023. Source: @tourismdata. The Australian international borders closed in March 2020 and officially reopened in February 2022."
#| fig-height: 3
tar_read(tourism_plot1)
```

@@

```{r}
#| label: fig-tsol1-plot
#| fig-height: 8
#| fig-pos: "!b"
#| fig-cap: "**Highly adaptive models**. Each panel shows one year of monthly forecasts from ETS and ARIMA models. Neither works particularly well for disruptions of this magnitude."
tar_read(tsol1_plot)
```

@@

```{r}
#| label: fig-tsol2-plot
#| fig-height: 8
#| fig-pos: "!b"
#| fig-cap: "**Intervention model**. Each panel shows one year of monthly forecasts from an intervention model containing a level shift from March 2020 to October 2022, and a ramp from October 2021 to October 2022. After October 2022, all intervention variables are set to zero."
tar_read(tsol2_plot)
```

@@

```{r}
#| label: fig-tsol3-plot
#| fig-height: 8
#| fig-pos: "!b"
#| fig-cap: "**Set to missing**. Each panel shows one year of monthly forecasts from an ARIMA model. The period from March 2020 to October 2022 is set to missing. So the first three years of forecasts show what might have been. The fourth set of forecasts uses observations from November and December 2022, and so the forecasts have been adjusted downwards."
tar_read(tsol3_plot)
```

@@

```{r}
#| label: fig-tsol4-plot
#| fig-height: 8
#| fig-pos: "!b"
#| fig-cap: "**Estimate what might have been**. Each panel shows one year of monthly forecasts from ETS and ARIMA models. We have replaced the observations from March 2020 to October 2022 with the average of the same month in the three years prior to March 2020."
tar_read(tsol4_plot)
```

@@

@@

```{r}
#| label: fig-walkers
#| fig-height: 3.7
#| fig-cap: "Average pedestrian traffic in Melbourne, Australia, from 2018 to 2023, based on automated hourly counts from sensors around the city. The counts are averaged over `r ngoodsensors` sensors, each of which had no more than 30 days of missing data. The gray shaded regions indicate periods of lockdown due to COVID-19. Source: @pedestrians"
tar_read(walkers_plot)
```

@@

@@

```{r}
#| label: fig-walkers-plot1
#| fig-height: 3
#| fig-cap: "**Highly adaptive model**: Rolling 7-day forecasts of daily pedestrian data, made each week, using ARIMA models. The forecasts do not take into account the lockdown periods, but the ARIMA models are largely able to adjust to the changing level of the series."
tar_read(walkers_plot1)
```

```{r}
#| label: fig-walkers-plot2
#| fig-height: 3
#| fig-cap: "**Intervention model**: Rolling 7-day forecasts of daily pedestrian data, made each week, using ARIMA models with a dummy variable indicating when the lockdown periods occurred. Many people were in self-imposed lockdown before the start of the first lockdown period, and were reluctant to return to the city after it ended, resulting in poor forecasts on either side of the first lockdown. The model was able to adjust to the changing level of the series during subsequent lockdowns."
tar_read(walkers_plot2)
```

```{r}
#| label: fig-walkers-plot3
#| fig-height: 3
#| fig-cap: "**Set to missing**: Rolling 7-day forecasts of daily pedestrian data, made each week, using ARIMA models, where the lockdown periods are set to missing. The ARIMA models use the remaining data to produce forecasts. Thus, the forecasts are based on only previous non-lockdown data. Naturally, the forecasts during the lockdown periods do not reflect reality, but they can be interpreted as what might have been."
tar_read(walkers_plot3)
```

```{r}
#| label: fig-walkers-plot4
#| fig-height: 3
#| fig-cap: "**Estimate what might have been**: Rolling 7-day forecasts of daily pedestrian data, made each week. The lockdown periods are replaced with estimated counts based on an ARIMA model applied to the remaining data. Then a new ARIMA model is fitted to each data set, and used to produce forecasts. As a result, the forecasts during the lockdown periods are not true forecasts, as they use non-lockdown data from the future."
tar_read(walkers_plot4)
```

```{r}
#| label: fig-walkers-ensemble-plot
#| fig-height: 3
#| fig-cap: "**Ensemble forecasts** combining the first two solutions shown in Figures [-@fig-walkers-plot1] and [-@fig-walkers-plot2]. Other than having some difficulty around the period of the first lockdown, the results are relatively good."
tar_read(walkers_ensemble_plot)
```

@@

@@

\textcolor{teal}{When forecasting interrupted time series data, exogenous variables associated with specific events should be identified and incorporated where possible, regardless of which approach is employed, since they may improve forecasting model performance.}

@@

We thank Dr Stephan Kolassa and several referees for their thoughts and feedback on the first version of our paper. These comments have helped us to refine and improve our work.
